---
title: "Project1_Math"
output: html_document
---

# Data Loading and Initial Exploration

```{r}
# Load required libraries
library(igraph)

# Read your data
edges <- read.csv("git_web_ml/musae_git_edges.csv")
features <- jsonlite::fromJSON("git_web_ml/musae_git_features.json")
target <- read.csv("git_web_ml/musae_git_target.csv")
```


## Edges Exploration
```{r}
names(edges)
sapply(edges, class)
head(edges)
tail(edges)
summary(edges)
str(edges)
dim(edges)
class(edges)
head(unique(edges), n = 10)
# table is way too big to display
#head(table(edges), n = 10)
hist(edges$id_1)
boxplot(edges$id_1)
plot(edges$id_1)
```

## Features Exploration
```{r}
head(names(features), n = 10)
head(sapply(features, class), n =10)
head(features)
tail(features)
#summary(features)
head(str(features), n = 10)
dim(features)
class(features)
head(unique(features), n = 10)
# table is way too big to display
#head(table(edges), n = 10)
all_features <- unlist(features)

# Create a histogram of the flattened numeric vector
hist(all_features, 
     main = "Histogram of Features", 
     xlab = "Feature Values", 
     col = "skyblue", 
     border = "white")
boxplot(all_features)
plot(all_features)

```

## Target Exploration
```{r}
head(names(target), n = 10)
head(sapply(target, class), n =10)
head(target)
tail(target)
#summary(features)
head(str(target), n = 10)
dim(target)
class(target)
head(unique(target), n = 10)
# table is way too big to display
#head(table(edges), n = 10)
# all_targets <- unlist(target)
# 
# # Create a histogram of the flattened numeric vector
hist(target$ml_target, 
     main = "Histogram of ML Targets", 
     xlab = "Target Values", 
     col = "lightgreen", 
     border = "white", 
     breaks = 2)  # Set breaks to 2 since there are only two classes (0 and 1)
boxplot(target$ml_target)
plot(target$ml_target)
```


# Plot Nodes
```{r}

# # Create graph from edges
# g <- graph_from_data_frame(edges, directed = FALSE)
# 
# # Simple plot
# plot(g,
#      vertex.size = 1,          # Node size
#      vertex.label = NA,        # Hide node labels
#      vertex.color = "skyblue", # Node color
#      edge.color = "gray90",    # Edge color
#      layout = layout_with_fr)  # Fruchterman-Reingold layout
```

``` {r}
# # Identify key nodes (e.g., high-degree hubs)
# deg <- degree(g)
# top_nodes <- names(sort(deg, decreasing = TRUE)[1:5])  # Top 5 nodes
# 
# # Create ego networks around these hubs
# subg <- make_ego_graph(g, order = 1, nodes = top_nodes) %>%
#   do.call(what = union)
# 
# # Plot focused subnetwork
# plot(subg,
#      vertex.size = 1,
#      vertex.label = NA,
#      vertex.color = adjustcolor("purple", alpha = 0.3),
#      edge.alpha = 0.05,
#      layout = layout_with_drl)
```

# Visualize the data and find the following parameter associated with the graph.
```{r}

# Create graph from edges
g <- graph_from_data_frame(edges, directed = FALSE)

# Order (number of nodes)
order <- vcount(g)
cat("Order of the graph (number of nodes):", order, "\n")

# Size (number of edges)
size <- ecount(g)
cat("Size of the graph (number of edges):", size, "\n")

# Degree sequence
degree_sequence <- sort(degree(g), decreasing = TRUE)

hist(degree_sequence,
     main = "Histogram of Degree Sequence (Log Scale)",
     xlab = "Degree (Log Scale)",
     ylab = "Frequency",
     col = "skyblue",
     border = "black",
     breaks = "FD", # Freedman-Diaconis rule for optimal bin width
     xlim = c(min(degree_sequence), max(degree_sequence)),
     log = "x") # Key: Set the x-axis to logarithmic scale


max_degree <- max(degree(g))
cat("Maximum degree of the graph:", max_degree, "\n")

# Minimum degree
min_degree <- min(degree(g))
cat("Minimum degree of the graph:", min_degree, "\n")

max_degree_node <- which.max(degree(g))

# Create ego network for the node with maximum degree
ego_network <- make_ego_graph(g, order = 1, nodes = max_degree_node)[[1]]

# Plot the ego network
plot(ego_network,
     vertex.size = 5,
     vertex.label = NA,
     vertex.color = "skyblue",
     edge.color = "gray90",
     main = paste("Ego Network of Node", max_degree_node))


# Vertex connectivity
vertex_connectivity <- vertex_connectivity(g)
cat("Vertex connectivity of the graph:", vertex_connectivity, "\n")

# Edge connectivity
edge_connectivity <- edge_connectivity(g)
cat("Edge connectivity of the graph:", edge_connectivity, "\n")

# Minimum cutset (vertex cutset)
min_cutset <- min_cut(g, value.only = FALSE)
cat("Minimum vertex cutset:", min_cutset$cut, "\n")

# Minimum edge cutset
min_edge_cutset <- min_cut(g, value.only = FALSE, capacity = NULL)
cat("Minimum edge cutset:", min_edge_cutset$cut, "\n")
