---
title: "Project 2"
author: "Logan Bolton"
output: pdf_document
date: "2025-02-08"
---
_Acknowledgement:_ This code was created through the repurposing of code found in the lecture notes and through collaboration with Claude 3.5 Sonnet and o3-mini. I have very little experience with R, so these AI tools were very helpful for me while fixing errors and determining the correct syntax to plot graphs.

# Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("/Users/log/Github/Spring2025Classes/social_networks/project2")
D <- read.csv("congress_network/congress.edgelist")
library(igraph) 
library(stringr)
library(ggplot2)
library(jsonlite)
```

```{r}
# print("Data structure:")
# str(D)
# print("First few rows:") 
# head(D)
```

```{r}


# Format the data
edges_df <- data.frame(
  from = as.numeric(sub("^(\\d+).*", "\\1", D$X0.4...weight...0.002105263157894737.)),
  to = as.numeric(sub("^\\d+\\s+(\\d+).*", "\\1", D$X0.4...weight...0.002105263157894737.)),
  weight = as.numeric(sub(".*'weight':\\s*([0-9.]+).*", "\\1", D$X0.4...weight...0.002105263157894737.))
)

# Create the graph
g <- graph_from_data_frame(edges_df, directed = TRUE)
```

# Graph Characteristics
## Network Understanding
```{r Network Understanding}
print("Network Order (number of vertices):")
vcount(g)

# Network size (number of edges)
print("Network Size (number of edges):")
ecount(g)

# Network density
print("Network Density:")
edge_density(g)

# Check strong connectivity
components <- components(g, mode="strong")
cat("\nNumber of strongly connected components:", components$no, "\n")
cat("Size of largest strongly connected component:", max(components$csize), "\n")

```

## Degree Distribution
```{r degree distribution}
# Calculate different degree measures
in_deg <- degree(g, mode="in")
out_deg <- degree(g, mode="out")
total_deg <- degree(g, mode="total")

par ( mfrow = c (1 ,2))
hist ( in_deg , main = " In - Degree Distribution " ,
xlab = " In - Degree " , ylab = " Frequency " , col = " lightblue " )
hist ( out_deg , main = " Out - Degree Distribution " ,
xlab = " Out - Degree " , ylab = " Frequency " , col = " lightgreen " )
```

## PageRank
```{r PageRank}
page_rank <- page_rank (g , weights = E ( g ) $weight , directed = TRUE ) $vector
```
## Hub and Authority Scores
_Note:_ The depreciation warning about 'hub_score' and 'authority_score' appears to not actually be true.
``` {r}
hub_scores <- hub_score(g , scale = TRUE ) $vector # Hub scores
auth_scores <- authority_score(g , scale = TRUE ) $vector # Authority scores
```
## Closeness Centrality
The graph is NOT strongly connected. Therefore, we should use harmonic closeness
``` {r}
harmonic_close <- harmonic_centrality (g , weights = E ( g ) $weight , mode = "out" )
closeness_scores <- closeness ( g , mode = "out" , normalized = TRUE )
is.connected(g, mode = "strong")

```

## Betweenness Centrality 
``` {r}
betweenness <- betweenness (g , weights = E ( g ) $weight , directed = TRUE , normalized = TRUE )

```



# Nodes
``` {r}
get_top_10 <- function (metric, metric_name) {
  top_indices <- order ( metric , decreasing = TRUE )[1:10]
  data.frame (
  Metric = rep(metric_name, 10) ,
  Node = top_indices ,
  Value = round ( metric [ top_indices ] , 4)
  )
}

top_nodes <- rbind (
  get_top_10 ( in_deg , "In-Degree" ) ,
  get_top_10 ( out_deg , "Out-Degree" ) ,
  get_top_10 ( page_rank , "PageRank" ) ,
  get_top_10 ( harmonic_close , "Harmonic Closeness" ) ,
  get_top_10 ( betweenness , "Betweenness" ) ,
  get_top_10 ( hub_scores, "Hub Scores") ,
  get_top_10 ( auth_scores, "Authority Scores")
)
# Print results
print ( " Top 10 nodes by different centrality measures : " )
print ( top_nodes )
```
```{r}
node_freq <- table(top_nodes$Node)

# Convert to dataframe and sort
node_freq_df <- data.frame(
  Node = names(node_freq),
  Frequency = as.numeric(node_freq)
)
node_freq_df <- node_freq_df[order(-node_freq_df$Frequency), ]

# Create the plot using base R
barplot(node_freq_df$Frequency[1:10], 
        names.arg = node_freq_df$Node[1:10],
        main = "Top 10 Most Frequent Nodes",
        xlab = "Node ID",
        ylab = "Frequency",
        las = 2)  # Rotate x-axis labels

print(node_freq_df$Node[1:10])
```

```{r}
library(ggplot2)
library(dplyr)

# 1. Degree Distributions (In-Degree and Out-Degree)
degree_plot <- ggplot(top_nodes[top_nodes$Metric %in% c("In-Degree", "Out-Degree"), ],
                      aes(x = reorder(as.factor(Node), -Value), y = Value, fill = Metric)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "In-Degree vs Out-Degree Distribution",
       x = "Node",
       y = "Degree Value") +
  scale_fill_brewer(palette = "Set1")

# 2. PageRank Distribution
pagerank_plot <- ggplot(top_nodes[top_nodes$Metric == "PageRank", ],
                        aes(x = reorder(as.factor(Node), -Value), y = Value)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "PageRank Distribution",
       x = "Node",
       y = "PageRank Value")

# 3. Harmonic Closeness Distribution
closeness_plot <- ggplot(top_nodes[top_nodes$Metric == "Harmonic Closeness", ],
                         aes(x = reorder(as.factor(Node), -Value), y = Value)) +
  geom_bar(stat = "identity", fill = "darkgreen") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Harmonic Closeness Distribution",
       x = "Node",
       y = "Harmonic Closeness Value")

# 4. Betweenness Distribution
betweenness_plot <- ggplot(top_nodes[top_nodes$Metric == "Betweenness", ],
                           aes(x = reorder(as.factor(Node), -Value), y = Value)) +
  geom_bar(stat = "identity", fill = "orange") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Betweenness Distribution",
       x = "Node",
       y = "Betweenness Value")

# 5. Hub and Authority Scores Distribution
hub_auth_plot <- ggplot(top_nodes[top_nodes$Metric %in% c("Hub Scores", "Authority Scores"), ],
                        aes(x = reorder(as.factor(Node), -Value), y = Value, fill = Metric)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Hub and Authority Scores Distribution",
       x = "Node",
       y = "Score Value") +
  scale_fill_brewer(palette = "Set2")

# 6. Boxplot of All Metrics (Normalized)
# Normalize the Value for each metric so they can be compared on the same scale
top_nodes_normalized <- top_nodes %>%
  group_by(Metric) %>%
  mutate(NormalizedValue = (Value - min(Value)) / (max(Value) - min(Value))) %>%
  ungroup()

boxplot_all <- ggplot(top_nodes_normalized, aes(x = Metric, y = NormalizedValue, fill = Metric)) +
  geom_boxplot() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Distribution of Normalized Centrality Measures",
       x = "Metric",
       y = "Normalized Value") +
  scale_fill_brewer(palette = "Set3")

# Print all plots
print(degree_plot)
print(pagerank_plot)
print(closeness_plot)
print(betweenness_plot)
print(hub_auth_plot)
print(boxplot_all)
```

## Notable Nodes Scratchpaper  
- Degree
  - 163, 212
- Page Rank
  - 212, 205
- Harmonic Closeness
  - 25, 93
- Betweenness Distribution
  - 71, 212, 25
- Hub
  - 76, 232
- Authority
  - 140, 212

- Frequent Across all Metrics
  - "212" "71"  "163" "25"  "39"  "72"  "151" "172" "205" "236"
  
## Chosen Nodes
- 212
  - Most frequent across all
  - High Authority
  - Great page rank
- 140 
  - High Authority
- 71
  - Frequent across benchmarks - 2nd most
  - Very high betweenness
- 25
  - Extremely high closeness
- 39
  - Good betweenness and harmonic closeness
  
```{r actual names}
json_data <- fromJSON("congress_network/congress_network_data.json")

usernameList <- json_data$usernameList
usernames <- usernameList[[1]]

chosen_nodes <- c(212, 140, 71, 25, 39)
chosen_names <- usernames[chosen_nodes]

# Print the chosen node names
print("Names of Most Influential Politicians:")
print(chosen_names)
```

# Justification
The nodes I chose were 212, 71, 25, 140 and 39.

One way to determine the most central nodes is to see which nodes are consistently high across multiple different measurements. Node 212 (RepGallagher) stands out as the most consistently central figure across multiple measurements, with exceptionally high authority and PageRank scores, suggesting that he not only receives attention from other influential politicians but also serves as a crucial node in the flow of political information. His high scores across different centrality measures indicate that his influence is multifaceted rather than limited to a single type of network interaction.  71 (RepCardenas) also appears frequently across this collection of measurements while also maintaining a very high betweenness score. This high betweenness score means that RepCardenas acts like a bridge between different political subsets.  25 (SenSanders) has a very high harmonic closeness that stands out amongst other nodes. This harmonic score means that SenSanders is very well connected and has large amounts of politicians in his local network. 140 (SenDuckworth) has a high authority meaning that he is followed by many powerful politicians. 39 (SenJohnHoeven) has strong betweenness and harmonic closeness scores.

