---
title: "Project 5"
output: pdf_document
author: "Logan Bolton"
date: "2025-02-23"
---

_Acknowledgement:_ This code was created through the repurposing of code found in the lecture notes and through collaboration with ChatGPT-4o by OpenAI. AI tools were very helpful for me while fixing errors and determining the correct syntax to plot graphs.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library('igraph')
library(poweRlaw)

```



```{r load_dataset}
url1 <- "https://raw.githubusercontent.com/JeffreyAlanSmith/Integrated_Network_Science/master/data/affiliations_1996.txt"
url2 <- "https://raw.githubusercontent.com/JeffreyAlanSmith/Integrated_Network_Science/master/data/affiliations_1996.txt"

affiliations96 <- read.delim(file = url1, check.names = FALSE)
affiliations97 <- read.delim(file = url2, check.names = FALSE)
dim(affiliations96)
```

# 1 - 1996 Dataset
## a - Which student clubs serve to integrate the school and which are more peripheral?
```{r}
G96 <- graph_from_incidence_matrix(as.matrix(affiliations96))
# Split nodes into types
V(G96)$type # TRUE = club, FALSE = student
clubs <- V(G96)[V(G96)$type == TRUE]

# Degree of each club (number of students in each club)
club_degrees <- degree(G96, v = clubs)

# Visualize distribution
hist(club_degrees, main = "Club Degree Distribution", xlab = "Number of Members")

# Identify top & bottom
head(sort(club_degrees, decreasing = TRUE), 10)  # Most connected (integrating)
head(sort(club_degrees, decreasing = FALSE), 10) # Least connected (peripheral)

```
## b - Which student clubs tend to share members at high rates?

```{r}
projections <- bipartite_projection(G96)
club_graph <- projections$proj2  # This is the club-club graph

# Weight = number of shared members
E(club_graph)$weight

# Top pairs with highest shared members
top_pairs <- sort(E(club_graph)$weight, decreasing = TRUE)
head(top_pairs, 10)
```
## c - What is the shared feature, or theme, that brings these clubs together in a cluster?


```{r}
club_comm <- cluster_louvain(club_graph)
plot(club_comm, club_graph, vertex.label = NA, main = "Communities of Clubs")

# View membership
membership(club_comm)
```

# 2 - 1997 Dataset

## a - What is the order, size, and density of G?
```{r}
G <- graph_from_incidence_matrix(as.matrix(affiliations97))

order_G <- gorder(G)
size_G <- gsize(G)
density_G <- edge_density(G)

cat("Order (number of vertices):", order_G, "\n")
cat("Size (number of edges):", size_G, "\n")
cat("Density:", density_G, "\n")
```

## b - Is the network G connected? If not, what fraction of vertices belong to the largest connected component? If the network is not connected, consider only the largest component H for the remaining questions.
```{r}
is_connected(G)
cat("The graph is NOT connected\n\n")

comp <- components(G)

# Size of largest component
largest_comp_size <- max(comp$csize)
fraction_largest <- largest_comp_size / gorder(G)

# Create subgraph H
H <- induced_subgraph(G, which(comp$membership == which.max(comp$csize)))

cat("Percent of vertices belonging to largest component: ", fraction_largest, "\n")
```

## c - What is the average path length of H?
```{r}
average_path_length_H <- average.path.length(H, directed = FALSE)
cat("Average path length: ", average_path_length_H)
```

## d - Is H scale-free? Provide statistical evidence (e.g., by examining the degree distribution and fitting a power-law distribution)

```{r}
deg <- degree(H)
hist(deg, breaks=50, main="Degree Distribution", xlab="Degree")

# Fit power law
fit <- fit_power_law(deg)
fit
```
## e - What is the fraction of edges that are attached to the top 10% of high-degree vertices?

```{r}
m <- displ$new(deg)
est <- estimate_xmin(m)
m$setXmin(est)
plot(m)
lines(m, col="red")
```
## f - What distributions do the following centrality measures follow:

```{r}
# Eigenvector
eigen_centrality <- eigen_centrality(H)$vector
hist(eigen_centrality, main="Eigenvector Centrality", xlab="Centrality")

# Betweenness
betw <- betweenness(H)
hist(betw, main="Betweenness Centrality", xlab="Betweenness")

# Closeness
closeness_centrality <- closeness(H)
hist(closeness_centrality, main="Closeness Centrality", xlab="Closeness")

```

## g - How does the clustering coefficient of vertices change with vertex degrees?
# FIX THIS
```{r}
clustering <- transitivity(H, type = "local", isolates = "zero")
plot(deg, clustering, xlab="Degree", ylab="Clustering Coefficient", main="Clustering vs Degree")

```
## h - Does H exhibit assortative mixing in terms of vertex degrees? Provide the assortativity coefficient and interpret its value.
```{r}
assortativity_degree <- assortativity_degree(H)
assortativity_degree

```

